A pipeline to analyse time-course gene expression data

### 2019-08-13 scripts/manuscript.Rmd
More fixes to the code
Bunch of fixes and additions to the manuscript

- Add legends to the plots that don't have.
- Currently have two functions "plot_genes", "plot_centroids", that do exactly
    the same thing. Find a better name for both of them
- switch to camel case function names
- Remove the namespace call
- Make clear that it applies to RNASeq and how to call the different functions
    to RNASeq
- Check that clustering can deal with count data.

@@

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression between distinct groups, there is also a
long history of time-course expression studies. Such studies compare the  gene expression
across time by measuring mRNA levels from samples collected at different timepoints [^1]. Such time-course studies can vary from measuring a few distinct time
points, to sampling ten to twenty time points. These longer time series are
particularly interested in investigating development over time. More recently,
a new variety of time course studies have come from single-cell sequencing experiments**EAP: need citation** which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
value that is not know but estimated from the data as its "pseudo-time."

@@

```{r schema, echo=FALSE, fig.width=3, fig.cap="Workflow for analyzing time-course datasets."}

@@

We then plot the two first PC components, and color each sample by (1) its
condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
**EAP: The colors for timepoint appear to circle around, i.e. the first and last are very similar colors**

```{r pca_plots, fig.width=6, fig.height=6}
par(mfrow=c(2, 2))
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Group", line=0.5 ,font=2,adj=0.5)
legend(x=250,y=25,legend=names(ann_colors$Group), fill=ann_colors$Group,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
legend(x=250,y=-15,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
```

@@

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points 
[@ritchie:limma, @robinson:edgeR, @love:moderated] . For long time-course datasets,
however, a separate test for each time-point results in creating many
different tests, for example one for every time point, the results of which
are difficult to integrate. We find in practice that the global analysis
simplifies analysis and of longer time courses data, with per-time point
analysis reserved for particularly interesting comparisons of individual
time-points.

We note that we could have time course data on either a single condition or
time course data on multiple conditions (such as the influenza dataset we are
considering), which will alter slightly the types of questions we are
interested in, but the two basic approaches remains the same. In what follows,
we will focus on the situation where we have multiple conditions. **EAP: Any
reason one and not the other. Maybe should change this sentence**

@@

@@

par(mfrow=c(2, 2))
timely_barplot(K_weekly_de_analysis, time, title="Kawasaki", las=3)
timely_barplot(C_weekly_de_analysis, time, title="California", las=3)
timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose", las=3)
timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose", las=3)
```

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
wh<-which(splines_model$meta$Group %in% c("K","M"))
plot_genes(data[exampleGenes,],splines_model,colors=ann_colors$Group)
```

**EAP: It would be nice to be able to subset in `plot_centroids/plot_genes` to only some of the groups**

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_genes(data[top_DE_genes_pval, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:10])
plot_genes(data[top_DE_genes_lfc, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

We then use the `plot_centroids` function to visualize the centroids obtained
with the splines k-means model.

```{r fig.height=6, fig.width=6}
plot_centroids(kmeans_clusters$centroids, splines_model,
    colors=ann_colors$Group,
    smooth=TRUE)
```

>>>>

- Add legends to the plots that don't have.
- Make clear that it applies to RNASeq and how to call the different functions
    to RNASeq
- Check that clustering can deal with count data.

@@

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression between distinct groups, there is also a
long history of time-course expression studies. Such studies compare the 
gene expression
across time by measuring mRNA levels from samples collected at different 
timepoints [^1]. Such time-course studies can vary from measuring a few 
distinct time
points, to sampling ten to twenty time points. These longer time series are
particularly interested in investigating development over time. More recently,
a new variety of time course studies have come from single-cell sequencing experiments
[@shalek:single-cell; @habib:div-seq; @trapnell:dynamics]
which can sequencing single cells which are at different stages of development; in this case,
the time point is the stage of the cell in the process of development -- a
value that is not know but estimated from the data as its "pseudo-time."

@@

```{r schema, echo=FALSE, fig.width=1, fig.cap="Workflow for analyzing time-course datasets."}

@@

We then plot the two first PC components, and color each sample by (1) its
condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.

```{r pca_plots, fig.width=6, fig.height=6}
par(mfrow=c(2, 2))
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275, text="Colored by Group", line=0.5, font=2, adj=0.5)
legend(x=250, y=25, legend=names(ann_colors$Group),
       fill=ann_colors$Group, bty="n", xpd=NA, yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Group[meta$Group],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
par(mar=c(4.5, 4.5, 1.5, 1.5))
plot(
    pca_data$x[, c("PC1","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
mtext(at=275, text="Colored by Timepoint", line=0.5, font=2, adj=0.5)
legend(x=250, y=-15, legend=names(ann_colors$Timepoint),
       fill=ann_colors$Timepoint, bty="n", xpd=NA, yjust=0.5)
par(mar=c(4.5, 5.5, 1.5, .1))
plot(
    pca_data$x[, c("PC3","PC2")],
    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
    ylab="", bty="l", yaxt="n")
axis(2, labels=FALSE)
```

@@

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points 
[@ritchie:limma; @robinson:edgeR; @love:moderated] . For long time-course datasets,
however, a separate test for each time-point results in creating many
different tests, for example one for every time point, the results of which
are difficult to integrate. We find in practice that the global analysis
simplifies analysis and of longer time courses data, with per-time point
analysis reserved for particularly interesting comparisons of individual
time-points.

We note that we could have time course data on either a single condition (to
identify genes changing over time) or
time course data on multiple conditions (such as the influenza dataset we are
considering), which will alter slightly the types of questions we are
interested in. 
In what follows, we will focus on the situation
where we have multiple conditions. **EAP: Any reason one and not the other.
Maybe should change this sentence**

@@

**NV: I'd rather not include these types of functions that are hard to test
and clutter the namespace. In order to provide reasonable visualization, they
also often need to be tweaked specfically to match the data (here, the ylim is
set specifically to the data**

@@

par(mfrow=c(2, 2))
time = names(time_colors)
timely_barplot(K_weekly_de_analysis, labels=time, title="Kawasaki", las=3)
timely_barplot(C_weekly_de_analysis, labels=time, title="California", las=3)
timely_barplot(weekly_de_analysis, labels=time, title="Vietnam low-dose", las=3)
timely_barplot(VH_weekly_de_analysis, labels=time, title="Vietnam high-dose", las=3)
```

@@

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
wh<-which(splines_model$meta$Group %in% c("K","M"))
plot_splines_data(data[exampleGenes,], splines_model, colors=ann_colors$Group)
```

**EAP: It would be nice to be able to subset in `plot_splines_data/plot_splines_data` to only some of the groups**

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

We then use the `plot_splines_data` function to visualize the centroids obtained
with the splines k-means model.

```{r fig.height=6, fig.width=6}
plot_splines_data(
    kmeans_clusters$centroids, splines_model,
    colors=ann_colors$Group,
    smooth=TRUE)
```
